Scalable gradients of Gaussian Process models with celerite

### 2018-01-25 paper/ms.tex
caption and acks

\section{Abstract}

@@

Gaussian Processes \citep[GPs][]{Rasmussen:2006} are a class of models used
extensively in the astrophysical literature to model stochastic processes.
The applications are broad-ranging and some example include the time domain
variability of astronomical sources \citep{Brewer:2009, Kelly:2014,
Haywood:2014, Rajpaul:2015, Foreman-Mackey:2017}, data-driven models of light
curves or stellar spectra \citep{Wang:2012, Luger:2016, Czekala:2017}, and the
cosmic microwave background \citep{Bond:1987,Wandelt:2003}.
In all of these applications, the calculation and optimization of the GP
marginalized likelihood function (here we follow the notation of
\citealt{Foreman-Mackey:2017})
\begin{eqnarray}\eqlabel{loglike}
\log \mathcal{L}(\bvec{\theta},\,\bvec{\alpha}) &=&
    -\frac{1}{2}\,\left[\bvec{y} - \bvec{\mu}_\bvec{\theta}\right]^\T\,
        {K_\bvec{\alpha}}^{-1}\,\left[\bvec{y}-\bvec{\mu}_\bvec{\theta}\right]
    -\frac{1}{2}\,\log\det K_\bvec{\alpha} + \mathrm{constant}
\end{eqnarray}
is of central importance and this can often be the computational bottleneck.
The details of this model are omitted here (more details can be found in
\citealt{Rasmussen:2006} and \citealt{Foreman-Mackey:2017}), but the key point
is that, for a dataset with $N$ data points, every evaluation of a GP model
requires computation of the inverse\footnote{Or, more specifically, a linear
operator that can left multiply a vector or a matrix by the inverse of this
matrix.} and log-determinant of the $N \times N$ covariance matrix.
The computational cost of these operations scales as $\mathcal{O}(N^3)$ in the
general case, but the \emph{celerite} method was recently introduced in the
astronomical literature to compute the GP likelihood for a class of
one-dimensional models with $\mathcal{O}(N)$ scaling
\citep{Foreman-Mackey:2017}.

@@

Following this reasoning and using the notation from an excellent review of
matrix gradients \citep{Giles:2008}, I present the reverse-mode gradients of
the \emph{celerite} method.
While not yet popular within astrophysics, ``reverse-mode accumulation'' of
gradients (also known as ``backpropagation'') has revolutionalized the field
of machine learning over the past few decades (TODO ADD CITATION).
The review \citep{Giles:2008} provides a thorough overview of these methods
and the interested reader is directed to that discussion for details and for
an explanation of the notation.

@@

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.9\textwidth]{figure.pdf}
\caption{%
    Scaling.
\figurelabel{figure}}
\end{center}
\end{figure}

%\acknowledgements
%It is a pleasure to thank
%  Patrick Cooper,
%  Boris Leistedt,
%  Bernhard Sch\"olkopf, and
%  Dun Wang
%for helping us understand all of this.

>>>>

@@

GPs \citep{Rasmussen:2006} are a class of models used
extensively in the astrophysical literature to model stochastic processes.
The applications are broad-ranging and some example include the time domain
variability of astronomical sources \citep{Brewer:2009, Kelly:2014,
Haywood:2014, Rajpaul:2015, Foreman-Mackey:2017}, data-driven models of light
curves or stellar spectra \citep{Wang:2012, Luger:2016, Czekala:2017}, and the
cosmic microwave background \citep{Bond:1987,Wandelt:2003}.
In all of these applications, the calculation and optimization of the GP
marginalized likelihood function (here we follow the notation of
\citealt{Foreman-Mackey:2017})
\begin{eqnarray}\eqlabel{loglike}
\log \mathcal{L}(\bvec{\theta},\,\bvec{\alpha}) &=&
    -\frac{1}{2}\,\left[\bvec{y} - \bvec{\mu}_\bvec{\theta}\right]^\T\,
        {K_\bvec{\alpha}}^{-1}\,\left[\bvec{y}-\bvec{\mu}_\bvec{\theta}\right]
    -\frac{1}{2}\,\log\det K_\bvec{\alpha} + \mathrm{constant}
\end{eqnarray}
is of central importance and this can often be the computational bottleneck.
The details of this model are omitted here (more details can be found in
\citealt{Rasmussen:2006} and \citealt{Foreman-Mackey:2017}), but the key point
is that, for a dataset with $N$ data points, every evaluation of a GP model
requires computation of the inverse\footnote{Or, more specifically, a linear
operator that can left multiply a vector or a matrix by the inverse of this
matrix.} and log-determinant of the $N \times N$ covariance matrix.
The computational cost of these operations scales as $\mathcal{O}(N^3)$ in the
general case, but the \emph{celerite} method was recently introduced in the
astronomical literature to compute the GP likelihood for a class of
one-dimensional models with $\mathcal{O}(N)$ scaling
\citep{Ambikasaran:2015, Foreman-Mackey:2017}.

@@

Following this reasoning and using the notation from an excellent review of
matrix gradients \citep{Giles:2008}, I present the reverse-mode gradients of
the \emph{celerite} method.
While not yet popular within astrophysics, ``reverse-mode accumulation'' of
gradients (also known as ``backpropagation'') has revolutionalized the field
of machine learning over the past few decades \citep[for
example][]{LeCun:2015}.
The review \citep{Giles:2008} provides a thorough overview of these methods
and the interested reader is directed to that discussion for details and for
an explanation of the notation.

@@

It is a pleasure to thank
Eric Agol,
Sivaram Ambikasaran, and
Victor Minden
for conversations that inspired this work.
A reference implementation of these algorithms, benchmarks, and tests can be
found at \url{https://github.com/dfm/celerite-grad}.

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{figure.pdf}
\caption{%
    The computational scaling of the algorithms prsented in this note.
    \emph{(top row)}: The computational cost of computing \eq{loglike} as a
    function of the number of datapoints ($N$; left) and the model complexity
    ($J$; right).
    \emph{(bottom row)}: The cost of computing the gradient of \eq{loglike}
    with respect to the vector $\bvec{a}$ and the matrices $U$, $V$, and $P$.
    In the left panels, each line corresponds to a different value of $J$ as
    indicated in the legend (with $J$ increasing from bottom to top).
    Similarly, in the right panels, the lines correspond to different values
    of $N$ increasing from bottom to top.
\figurelabel{figure}}
\end{center}
\end{figure}
